---
title: "Getting Started with rlistings"
date: "2022-10-26"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float:
          collapsed: false
vignette: >
    %\VignetteIndexEntry{Getting Started}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown: 
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

---------

This vignette shows the general purpose and basic functionality of the `rlistings` R package.  
The `rlistings` R package contains value formatting and ASCII rendering infrastructure for tables and listings useful for clinical trials and other statistical analysis. 
The core functionality is built on top of the [`formatters` package](https://insightsengineering.github.io/formatters/). 

Some of the key features currently available to customize listings created using the `rlistings` package include:

- Key columns
- Titles and footnotes

The index of all available `rlistings` functions can be found on the [`rlistings` website functions reference](https://insightsengineering.github.io/rlistings/reference/index.html).

The `rlistings` package is intended for use in creating simple one-dimensional listings. For construction of more complex tables see the [`rtables` package](https://roche.github.io/rtables/).

---------

## Building a Listing

With the basic framework provided in this package, a `data.frame` object can be easily converted into a listing using the `as_listing` function with several optional customizations available.

In the R code below we will give a basic example of how to create an `rlistings` listing from a pre-processed data frame.

We first load in the `rlistings` package.

```{r}
library(rlistings)
```

For the purpose of this example we will use the dummy ADAE dataset provided within the `formatters` package as our data frame, which consists of 48 columns of adverse event patient data.

```{r}
adae <- formatters::ex_adae
```

Now we will create our listing.

The `df` parameter sets our `data.frame` object. The `cols` argument takes a vector of names of any columns taken from the data frame that should be included in the listing. Column headers are set by the `label` attribute of each given variable. If there is no label associated with a given variable then the variable name will be taken as a header instead.

Since the ADAE dataset consists of 1934 rows in total, we will use the `head` function to print only the first 15 rows of the listing.

```{r}
lsting <- as_listing(
  df = adae,
  cols = c("USUBJID", "AGE", "SEX", "RACE", "ARM", "AEDECOD", "AESEV", "AETOXGR")
)

head(lsting, 15)
```

In the listing output above you can see that there are several rows associated with each patient, resulting in many instances of repeated values over several columns. This can be neatened by setting key columns with the `key_cols` argument.

## Key Columns

To tidy the listing and avoid repeatedly printing the same values in subsequent columns, the `key_cols` argument can take a vector of names of columns that should be treated as key columns in a listing. The data frame is sorted on the key columns and data repeated across all key columns for any subsequent rows will not be printed. You should note that any key columns set must also be included in the `cols` argument to be displayed in the listing. Below we implement key columns to clean up repeated patient-identifying information for improved readability.

```{r}
lsting <- as_listing(
  df = adae,
  cols = c("USUBJID", "AGE", "SEX", "RACE", "ARM", "AEDECOD", "AESEV", "AETOXGR"),
  key_cols = c("USUBJID", "AGE", "SEX", "RACE", "ARM")
)

head(lsting, 15)
```

## Titles and Footnotes

Additionally, an `rlistings` listing can be annotated with two types of header information (main title and subtitles) and two types of footer information (main footers and provenance footers). A single title can be set using the `main_title` argument, while one or more subtitles, main footers, and provenance footers can be set by the `subtitles`, `main_footer` and `prov_footer` arguments respectively. These are demonstrated in the following updated listing.

```{r}
lsting <- as_listing(
  df = adae,
  cols = c("USUBJID", "AGE", "SEX", "RACE", "ARM", "AEDECOD", "AESEV", "AETOXGR"),
  key_cols = c("USUBJID", "AGE", "SEX", "RACE", "ARM"),
  main_title = "Main Title",
  subtitles = c("Subtitle A", "Subtitle B"),
  main_footer = c("Main Footer A", "Main Footer B", "Main Footer C"),
  prov_footer = c("Provenance Footer A", "Provenance Footer B")
)

head(lsting, 15)
```

---------

## Summary

In this vignette you have learned how to implement the basic listing framework provided by the `rlistings` package to build a simple listing. You have also seen examples demonstrating how the optional parameters of the `as_listing` function can be set to customize and annotate your listings.

**For more information please explore the [rlistings website](https://insightsengineering.github.io/rlistings/).**
