---
title: "Getting Started with rlistings"
author: "Emily de la Rua and Gabriel Becker"
date: "2022-10-26"
output:
  rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Getting Started}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown: 
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette shows the general purpose and basic functionality of the `rlistings` R package.  

The `rlistings` R package contains value formatting and ASCII rendering infrastructure for tables and listings useful for clinical trials and other statistical analysis. 
The core functionality is built on top of the [`formatters` package](https://insightsengineering.github.io/formatters/latest-tag/).

Some of the key features currently available to customize listings created using the `rlistings` package include:

- Key columns
- Titles and footnotes

For information on listing column formatting see the [Column Formatting vignette](col_formatting.html).
To learn about listing pagination see the [Pagination vignette](pagination.html).

The index of all available `rlistings` functions can be found on the [rlistings website functions reference](https://insightsengineering.github.io/rlistings/main/reference/index.html).

The `rlistings` package is intended for use in creating simple one-dimensional listings. For construction of more complex tables see the [`rtables` package](https://insightsengineering.github.io/rtables/).

---------

## Building a Listing

With the basic framework provided in this package, a `data.frame`
object can be easily converted into a listing using the `as_listing`
function with several optional customizations available.

A listing, at its core, is a set of observation-level data which is to
be rendered with particular formatting but without any sort of
aggregation or further analysis. In practice, this translates to to a
classed `data.frame` (or `tbl_df`) object with a specialized print
method. This means that, unlike tables created with `rlistings`'
sibling package `rtables`, a listing object is fundamentally the
incoming `data.frame` with a few annotations attached to it.


In the R code below we will give a basic example of how to create an
`rlistings` listing from a pre-processed data frame.

We first load in the `rlistings` package.

```{r}
library(rlistings)
```

For the purpose of this example we will use the dummy ADAE dataset
provided within the `formatters` package as our data frame, which
consists of 48 columns of adverse event patient data, and one or more
rows per patient.

```{r}
adae <- ex_adae
```

Now we will create our listing.

The `df` parameter sets our `data.frame` object. The `disp_cols`
argument takes a vector of names of any columns taken from the data
frame that should be included in the listing. Column headers are set
by the `label` attribute of each given variable. If there is no label
associated with a given variable then the variable name will be taken
as a header instead. For this example we will choose 8 arbitrary
columns to display - 5 specific to the patient and 3 relating to the
adverse event.

Since the dataset consists of 1934 rows in total, we will use the
`head` function to print only the first 15 rows of the listing.

```{r}
lsting <- as_listing(
  df = adae,
  disp_cols = c("USUBJID", "AETOXGR", "ARM", "AGE", "SEX", "RACE", "AEDECOD", "AESEV"),
)

head(lsting, 15)
```

In the listing output above you can see that there are several rows
associated with each patient, resulting in many instances of repeated
values over several columns. This can cleaned up by setting key
columns with the `key_cols` argument.

We can also declare the set of (non-key) display columns by compliment,
via the `non_disp_col` argument. If specifies this argument accepts
names of columns which will non be displayed. All other non-key columns
are then displayed.


```{r}
lsting <- as_listing(
  df = adae,
  non_disp_cols = tail(names(adae), 8)
)
head(lsting, 15)
```

## Key Columns

Key columns act as contextual identifiers for observations. Their core
behavioral feature is that sequentially repeated values are not
displayed when they do not add information.

In practice, this means that each value of a key column is printed
only once per unique combination of values for all higher-priority
(i.e., to the left of it) key columns (per page). Locations where a
repeated value would have been printed within a key column for the
same higher-priority-key combination on the same page are rendered as
empty space.  Note, determination of which elements to display within
a key column at rendering is based on the underlying value; any
non-default formatting applied to the column has no effect on this
behavior.

The `key_cols` argument takes a vector of column names identifying the
key columns for the listing. A listing is always sorted by its key
columns (with order defining the sort precedence). Below we specify
trial arm and patient ID as key columns to improve readability.

```{r}
lsting <- as_listing(
  df = adae,
  disp_cols = c("ARM", "AGE", "SEX", "RACE", "AEDECOD", "AESEV"),
  key_cols = c("USUBJID", "AETOXGR")
)

head(lsting, 15)
```

## Titles and Footers

Additionally, an `rlistings` listing can be annotated with two types
of header information (main title and subtitles) and two types of
footer information (main footers and provenance footers). A single
title can be set using the `main_title` argument, while one or more
subtitles, main footers, and provenance footers can be set by the
`subtitles`, `main_footer` and `prov_footer` arguments
respectively. These are demonstrated in the following updated listing.

```{r}
lsting <- as_listing(
  df = adae,
  disp_cols = c("ARM", "AGE", "SEX", "RACE", "AEDECOD", "AESEV"),
  key_cols = c("USUBJID", "AETOXGR"),
  main_title = "Main Title",
  subtitles = c("Subtitle A", "Subtitle B"),
  main_footer = c("Main Footer A", "Main Footer B", "Main Footer C"),
  prov_footer = c("Provenance Footer A", "Provenance Footer B")
)

head(lsting, 15)
```

---------

## Summary

In this vignette you have learned how to implement the basic listing framework provided by the `rlistings` package to build a simple listing. You have also seen examples demonstrating how the optional parameters of the `as_listing` function can be set to customize and annotate your listings.

**For more information please explore the [rlistings website](https://insightsengineering.github.io/rlistings/main/).**
